Statistical performance analysis of neural mass models?

### 2013-06-13 manuscript/manuscript.tex
Draft of theory section

Not claiming anything about which model is the 'best', but simply what level of complexity is possible - how increasing complexity affects uncertainty.
\section{Background}

@@

The parameters of the neural masses not only define the population type, but also the behaviour exhibited by the model. For example, certain parameter combinations result in a model of a cortical column that will generate alpha wave type activity (normal activity) and, for another set of parameters, we create a model that will exhibit epileptic behaviour \cite{Wendling2002}. Therefore, we consider a family of models, which we define generally as 
\begin{align}\label{eq:NeuralMassModel}
    \dot{\mathbf{x}}(t) =& f_{\theta}\left(\mathbf{x}(t),\mathbf{u}(t)\right)\\
    y(t) =& \mathbf{C}\mathbf{x}(t) + e(t),
\end{align}
where $\mathbf{x}(t) \in \mathbb{R}^{n_x}$ is a state vector representing the postsynaptic membrane potentials generated by each population synapse and their time derivatives, $n_x$ is the number of states and $\mathbf{u}(t)$ represents the system input, which may be from afferent connections or other brain regions, exogenous input, or account for model inaccuracies. The function $f_{\theta}(\cdot)$ describes the dynamics, where $\theta \in \mathbb{R}^{n_{\theta}}$ determines the model type and the behaviour it exhibits. The iEEG is denoted by $y(t)$, $\mathbf{C}$ is the observation matrix, and $e(t)$ is the observation noise.

@@

Present 3 models?

\section{Theory / Performance Bounds}

Introduce recursive computation of bound and Monte Carlo approximation

\section{Methods?}

\section{Results}

@@

>>>>

Not claiming anything about which model is the 'best', but simply what level of complexity is possible - how increasing complexity affects uncertainty.
\section{Background}
\label{sec:background}

@@

The parameters of the neural masses not only define the population type, but also the behaviour exhibited by the model. For example, certain parameter combinations result in a model of a cortical column that will generate alpha wave type activity (normal activity) and, for another set of parameters, we create a model that will exhibit epileptic behaviour \cite{Wendling2002}. Therefore, we consider a family of models, which we define generally as 
\begin{align}\label{eq:NeuralMassModel}
    \dot{\vec{x}}(t) =& \vec f_{\theta}\left(\vec{x}(t),\vec{u}(t)\right)\\
    y(t) =& \mat{C}\vec{x}(t) + e(t),
\end{align}
where $\vec{x}(t) \in \mathbb{R}^{n_x}$ is a state vector representing the postsynaptic membrane potentials generated by each population synapse and their time derivatives, $n_x$ is the number of states and $\vec{u}(t)$ represents the system input, which may be from afferent connections or other brain regions, exogenous input, or account for model inaccuracies. The function $\vec f_{\theta}(\cdot)$ describes the dynamics, where $\theta \in \mathbb{R}^{n_{\theta}}$ determines the model type and the behaviour it exhibits. The iEEG is denoted by $y(t)$, $\vec{C}$ is the observation matrix, and $e(t)$ is the observation noise.

@@

\section{Theory}

The Bayesian Cramer-Rao bound (BCRB) can be used to examine the inherent uncertainty in a neural mass model and also assess the performance of different estimation algorithms.

INCLUDE TIME DEPENDENCE STRAIGHT AWAY

Central to the BCRB is the Bayes information matrix defined as
\begin{align}
	\mat J_k &= \mathbb E_{\vec x,\vec y}\left[ \nabla_{\vec x}\nabla_{\vec x}^T \log p(\vec y,\vec x) \right] 
\end{align}
The expectation is over the state, $\vec x$ and the measurements, $\vec y$ with respect to the joint distribution $p(\vec y,\vec x)$.

The BCRB is based on the Bayes information matrix, $J_{\rm B}$, which contains a data term and a prior term \cite{sdf},
\begin{align}
	J_{\rm B} = J_{\rm D} + J_{\rm P}
\end{align}
where 
\begin{align}
	J_{\rm D} &= \mathbb E_{\vec x,\vec y}\left[ \nabla_{\vec x}\nabla_{\vec x}^T \log \ell(\vec y|\vec x) \right] \\
	J_{\rm P} &= \mathbb E_{\vec x}\left[ \nabla_{\vec x}\nabla_{\vec x}^T \log p(\vec x) \right]
\end{align}
The data term, $J_{\rm D}$, depends on the likelihood, $\ell$, and the prior term, $J_{\rm P}$, depends on the prior distribution, $p$. The BCRB is ideally suited to filtering applications since the prior distribution captures the information up to the current time.

The BCRB gives a lower bound on the mean squared error
\begin{align}
	\mathsf{MSE}_k = \mathbb E_{\vec x_k,\vec y} [(\hat{\vec x}_k(\vec y) - \vec x_k)(\hat{\vec x}_k(\vec y) - \vec x_k)^T] \ge \mat J_{k}^{-1}
\end{align}
where $\hat{\vec x}_k$ is the estimated state at time $k$ given the measurements from time $1,\ldots,k$.

\subsection{Recursive Bayesian CRB}

It is necessary to develop a recursive formula for $\mat J_k$ for filtering applications to prevent the above computations becoming unwieldy. The popular recursion in \cite{Tichavsky1998} is not applicable to the neural field models described in Section~\ref{sec:background} since noise only enters the system through a single state and the resultant covariance matrix, $\mat Q$, is singular. The recursion in \cite{Bergman2001} is applicable to this class of models and can be summarised as 
\begin{align}
	\mat J_k = \left( \mat Q + \bar{\mat F}_{k-1} \mat J_{k-1}^{-1} \bar{\mat F}_{k-1}^T\right)^{-1} + \mat C^T \mat R^{-1} \mat C
\end{align}
where
\begin{align}
		\bar{\mat F}_k &= \mathbb E_{\vec x_k} \left[ \nabla_{\vec x_k}^T \vec f(\vec x_k)\right] 
		\label{eqn:F_matrix_expectation}
\end{align}
See Appendix~\ref{sec:derivation_recursion} for details. The expectation over the state $\vec x_k$ in \eqref{eqn:F_matrix_expectation} can rarely be computed in closed form so Monte Carlo techniques are employed. A large number of state realisations are generated at each time point and the expectation is approximated by an average over the realisations.

\section{Simulations}

@@

\appendix
\section{Derivation of Recursive Bayesian CRB}
\label{sec:derivation_recursion}

The recursion in \cite{Bergman2001} is
\begin{subequations}
\begin{align}
	\mat P_{k|k-1} &= \bar{\mat F}_k\mat P_{k-1|k-1}\bar{\mat F}_k + \bar{\mat G}_k\mat Q_k \bar{\mat G}_k^{-1} \\
		\mat P_{k|k} &= \left(\mat P_{k|k-1}^{-1} + \bar{\mat R}_{k-1}^{-1}\right)^{-1}
\end{align}
\end{subequations}
where
\begin{align}
	\bar{\mat F}_k &= \mathbb E \left[ \nabla_{\vec x_k}^T \tilde{\vec f_k}(\vec x_k,\vec v_k)\right] \\
	\bar{\mat R}_k^{-1} &= -\mathbb E\left[ \nabla_{\vec x_k}\nabla_{\vec x_k}^T \log p(\vec y_k|\vec x_k) \right] \\
	\bar{\mat G}_k &= \mathbb E \left[ \nabla_{\vec v_k}^T \tilde{\vec f_k}(\vec x_k,\vec v_k)\right] \\
	\mat Q_k^{-1} &= -\mathbb E \left[ \nabla_{\vec v_k}\nabla_{\vec v_k}^T \log p(\vec v_k) \right]
\end{align}
and $\mat P_0^{-1}$ is the inverse of the prior covariance. The model in \eqref{eqn:model_general} is represented by $\tilde{\vec f}(\vec x_k,\vec v_k) = \vec f(\vec x_k) + \vec v_k$. This simplifies the derivatives to
\begin{align}
	\nabla_{\vec x_k}^T \tilde{\vec f_k}(\vec x_k,\vec v_k) &= \mat F_k(\vec x_k) \\
	\nabla_{\vec v_k}^T \tilde{\vec f_k}(\vec x_k,\vec v_k) &= \mat I_{d\times d} \\
	\nabla_{\vec x_k}\nabla_{\vec x_k}^T \log p(\vec y_k|\vec x_k) &= -\mat H^T\mat R^{-1} \mat H 
\end{align}
and the corresponding matrices to	
\begin{align}
	\bar{\mat F}_k &= \mathbb E\left[ \mat F_k(\vec x_k)\right] \label{eqn:pcrb_term_F}\\
	\bar{\mat R}_k^{-1} &= \mat H^T\mat R^{-1} \mat H \\
	\bar{\mat G}_k &= \mat I_{d\times d} 
\end{align}
Therefore the final recursion is
\begin{align}
	\mat J_{k|k} = \left( \mat Q + \bar{\mat F}_{k-1} \mat J_{k-1}^{-1} \bar{\mat F}_{k-1}^T\right)^{-1} + \mat H_k^T \mat R^{-1} \mat H_k
\end{align}
